---
title: "R Notebook"
output: 
---

How are the number of releases and number of vignettes related for R packages?

https://juliasilge.com/blog/rstats-vignettes/

```{r}
library(tidyverse)
cran <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-03-15/cran.csv")
```

```{r}
cran |> filter(package == "tidytext")
glimpse(cran)
```

Compute the first release date, the number of releases, and the number of vignettes as of the most recent release. Also compute proportion of packages with no vignettes.
```{r}
vignette_counts <- cran |> 
  group_by(package) |> 
  summarise(
    release_date = first(date),
    releases = n(),
    vignettes = last(rnw) + last(rmd)
  )

mean(vignette_counts$vignettes < 1)
```

Plot number of vignettes in a histogram.

```{r}
vignette_counts |> 
  ggplot(aes(vignettes)) +
  geom_histogram(bins = 12) +
  scale_x_continuous(trans = scales::pseudo_log_trans(base = 10))
```

Just a few packages have a ton of vignettes.

```{r}
vignette_counts |> filter(vignettes > 20)
```

One more plot.

```{r}
vignette_counts |> 
  mutate(has_vignette = vignettes > 0) |> 
  ggplot(aes(has_vignette, releases, color = has_vignette, fill = has_vignette)) +
  geom_boxplot(size = 1.5, alpha = 0.2, show.legend = FALSE) +
  scale_y_log10() +
  coord_flip() +
  labs(x = "Has a vignette?")
```

Use Poisson regression to analyze this count data. Load the poissonreg package. The model says that packages with more releases have more vignettes. 

```{r}
library(tidymodels)
library(poissonreg)

poisson_wf <- workflow(vignettes ~ releases, poisson_reg())
fit(poisson_wf, data = vignette_counts)
```

A zero-inflated Poisson (ZIP) model mized two models, one that generates zeros and one that models counts with the Poisson distribution. There are two sets of covariates, one for count data and one for the probability of zeros.

```{r}
zip_spec <- poisson_reg() %>% set_engine("zeroinfl")
zip_wf <- workflow() %>%
  add_variables(outcomes = vignettes, predictors = releases) %>%
  add_model(zip_spec, formula = vignettes ~ releases | releases)

fit(zip_wf, data = vignette_counts)
```

